### 2014-06-02 writing/S'14 paper/abstract.tex
Incremental paper update.

\begin{abstract}
Supervised recursive neural network models (RNNs) for sentence meaning
have been successful in a wide array of sophisticated language tasks,
but it remains an open question whether they can achieve the same
results as compositional semantic grammars based in logical forms. We
address this question directly by using a logical grammar to generate
controlled data sets encoding the relationships (entailment, synonymy,
contradiction) between pairs of expressions and evaluating whether
each of two classes of neural model --- plain RNNs and recursive
neural tensor networks (RNTNs) --- can learn those relationships
correctly. Our first experiment confirms that both models can learn
the basic algebra of logical relations involved. Our second and third
experiments expand on this result to cover complex recursive
structures and sentences involving quantification. We find that the
plain RNN achieves only mixed results on the second and third
experiments, but that the stronger RNTN models generalized well in
every case.
\end{abstract}

>>>>

\begin{abstract}
Supervised recursive neural network models (RNNs) for sentence meaning
have been successful in a wide array of sophisticated language tasks,
but it remains an open question whether they can achieve the same
results as compositional semantic grammars based in logical forms. We
address this question directly by using a logical grammar to generate
controlled data sets encoding the relationships (entailment, synonymy,
contradiction) between pairs of expressions and evaluating whether
each of two classes of neural model --- plain RNNs and recursive
neural tensor networks (RNTNs) --- can learn those relationships
correctly. Our first experiment confirms that both models can learn
the basic algebra of logical relations involved. Our second and third
experiments expand on this result to cover complex recursive
structures and sentences involving quantification. We find that the
plain RNN achieves only mixed results on the first and second
experiments, but that the stronger RNTN models generalized well in
every case. % TODO: Update to reflect results
\end{abstract}

### 2014-06-02 writing/S'14 paper/methods.tex
Incremental paper update.

\section{Methods}

The model that I study is a recursive neural network model \citet{socher2011dynamic}, and is thus centered on a recursively applied composition function which is meant to mimic the recursive construction of meanings in formal models of semantics. In this scheme, pairs of words are merged into phrase representations by a function that maps from representations of length $2N$ to representations of length $N$. These phrase representations are then further merged with other words and phrases until the entire phrase or sentence being evaluated is represented in a single vector. This vector is then used as the input to a classifier and used in a supervised learning task.

Borrowing a model directly from the existing literature for this task is impossible since none has been proposed to detect asymmetric relations between phrases. Instead, I build a combination model, depicted in Figure \ref{sample-figure}. The two phrases being compared are built up separately on each side of the tree using the same composition function until they have each been reduced to single vectors. Then, the two phrase vectors are fed into a separate comparison layer that is meant to generate a feature vector capturing the relation between the two phrases. The output of this layer is then given to a softmax classifier, which in turn produces a hypothesized distribution over the seven relations.

For a composition layer, I experiment with both a plain neural network layer function (eq. \ref{rnn}) and the RNTN layer function proposed in \citet{chen2013learning} (eq. \ref{rntn}). A sigmoid nonlinearity (elementwise $\tanh$) is applied to the output of either layer function, following Socher et al.

 \label{rnn}\label{rntn} % TODO: Fix labeling
\begin{equation}
y_{RNN} = f(M [\vec{x}^{(l)}; \vec{x}^{(r)}] + b_i)\\
y_{RNTN}^i = f(\vec{x}^{(l)T} \mathbf{A^{[i]}} \vec{x}^{(r)} + \vec{B_{i,:}} [\vec{x}^{(l)}; \vec{x}^{(r)}] + c_i)
\end{equation}

@@

\textit{Model size:} The dimensionalities of the model were tuned extensively. The performance of the model is approximately optimal with the dimensionality of the word vectors set to 16, and the dimensionality of the feature vector produced by the comparison layer set to 45. % TODO: Later

\begin{figure}
\begin{center}
\Tree [.{\sc softmax classifier}  [.{\sc comparison layer} [.{\sc \ii{dog} vector from $V$} ] [.{\sc composition layer} [.{\sc composition layer} [.{\sc \ii{very} vector from $V$} ] [.{\sc \ii{big} vector from $V$} ] ] [.{\sc \ii{cat} vector from $V$} ] ] ] ]
\end{center}
\caption{The model structure used to compare \ii{dog} and \ii{(very big) cat}. \label{sample-figure}} 
\end{figure} % TODO: Create a more compact figure using something better than qtree.

>>>>

\section{Recursive neural network models}

The models that we study is are centered aronund a recursively applied composition function which is meant to mimic the recursive construction of meanings in formal semantics. In this scheme, pairs of words are merged into phrase representations by a function that maps from representations of length $2N$ to representations of length $N$. These phrase representations are then further merged with other words and phrases until the entire phrase or sentence being evaluated is represented in a single vector. This vector is then used as the input to a classifier and used in a supervised learning task.

Borrowing a model directly from the existing literature for this task is impossible since none has been proposed to detect asymmetric relations between phrases. Instead, we build a combination model, depicted in Figure \ref{sample-figure}. The two phrases being compared are built up separately on each side of the tree using the same composition function until they have each been reduced to single vectors. Then, the two phrase vectors are fed into a separate comparison layer that is meant to generate a feature vector capturing the relation between the two phrases. The output of this layer is then given to a softmax classifier, which in turn produces a hypothesized distribution over the seven relations.

For a composition layer, we experiment with both a plain neural network layer function (eq. \ref{rnn}) and the RNTN layer function proposed in \citet{chen2013learning} (eq. \ref{rntn}). A sigmoid nonlinearity (elementwise $\tanh$) is applied to the output of either layer function, following Socher et al.
\begin{gather} \label{rnn}
y_{RNN} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + b_i)\\ % TODO: Add column vectors?
\label{rntn}
y_{RNTN} = f(\vec{x}^{(l)T} \mathbf{A}^{[1...N]} \vec{x}^{(r)} + \mathbf{B} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{c})
\end{gather} % TODO: Explain third order tensor parameter?

@@

\begin{figure}
\begin{center}
% \Tree [.{\sc softmax classifier}  [.{\sc comparison layer} [.{\sc \ii{dog} vector from $V$} ] [.{\sc composition layer} [.{\sc composition layer} [.{\sc \ii{very} vector from $V$} ] [.{\sc \ii{big} vector from $V$} ] ] [.{\sc \ii{cat} vector from $V$} ] ] ] ]
\includegraphics[scale=0.36]{model.eps}
\end{center}
\caption{The model structure used to compare \ii{no dog} and \ii{(not all) dog}. \label{sample-figure}} 
\end{figure} % TODO: Create a more compact figure using something better than qtree.

### 2014-06-02 writing/S'14 paper/quantifiers.tex
Incremental paper update.

% Introduce the task

\subsection{Data}

@@

% TODO: Mention all seven relations seen, but some rare

\section{Experiments and results}

% TODO: Update with newer experiments

In the simplest experimental setting, which I label \textsc{all-split}, I randomly sample 85\% the data---making sure to sample 85\% of each of the individual datasets---train the model on that portion, and evaluate on the remaining 15\% of the data. This setting is meant to test whether the model is able to correctly generalize the individual reasoning patterns represented by each of the datasets.

Performance on this setting is perfect: the model quickly converges to 100\% accuracy on the test data, showing that it is capable of accurately learning to capture all of the reasoning patterns in the data. The remaining experimental settings serve to determine whether what is learned captures the underlying logical structure of the data in a way that allows it to accurately label unseen kinds of reasoning pattern. In each of them, I choose one of three arbitrarily chosen target datasets, all involving quantifier substitution, to test on. I then then hold out that dataset and---depending on the experimental setting---other similar datasets from the training data in an attempt to discover just how different a test example can be from anything seen in training and still be classified accurately. Table \ref{patterntable} shows what information is included in the training data for each of the four settings for one of the three target datasets.

\begin{table}\small\centering
\begin{tabular}{|l|l|}\hline
\textbf{Training configuration} & \textbf{Test accuracy}\\\hline
\textsc{all split}	&	TODO\% \\\hline
\textsc{pair no/no}	&	TODO\% \\
\textsc{pair two/less-than-two}	&	TODO\% \\
\textsc{pair not-all/not-most}	&	TODO\% \\
\textsc{pair all/some}	&	TODO\% \\\hline
\end{tabular}

@@

% TODO: Define experimental setups

\subsection{Discussion}

@@

The model learns to generalize well to novel data in most but not all of the training configurations. This inconsistent performance suggests that there is room to improve the optimization techniques used on the model, but the fact that it is able to perform well in these settings even some of the time suggests that the structure of the model is basically capable of learning meaning representations that support inference.

The results in the \textsc{all-split} condition show two important behaviors. The model is able to learn to identify the difference between two unseen sentences and consistently return the label that corresponds to that difference. In addition, the model can learn lexical relations from the training data, such as \ii{dog $\sqsubset$ animal} from \ii{(no dog) bark $\sqsupset$ (no animal) bark}, and it can then use these learned lexical relations to compute the relation for a sentence pair like \ii{(some dog) bark $\sqsubset$ (all animal) bark}. The results from the other three experimental settings show that the model is able to learn general purpose representations for quantifiers which, at least in many cases, allow it to perform inference when a crucial difference between two sentences---the  substitution of one specific quantifier for another---has not been seen. These results serve to confirm that the representations learned are capturing some of the underlying logic, rather than than just supporting the memorization of fixed reasoning patterns.

%  There has also been some research, most recently including \citet{grefenstette2013towards}, into models which represent quantifiers and other function words as higher order tensors rather than vectors, motivated in part by the hypothesis that their compositional behavior cannot be captured in strictly vector-based representatons. A successful inference engine that used those representations would not be a strict rebuttal of this assumption, but it could provide insight into precisely what circumstances would warrant the use of these richer tensor representation models in practice.  % TODO: Proof

These results leave open the question of how much information is minimally needed to learn general purpose representations for quantifiers in this setting. There are two lines of experimental work that could help to clarify this. Including longer sentences and constructions involving conjunctions (i.e. \ii{and, or}), transitive verbs (i.e. \ii{eats, kicks}) or other constructions in the training and test data could further test what kinds of behavior can be learned from a small training set, as could further experiments on this data involving even smaller training sets than those shown here or differently structured configurations of train and test sets.

%The representations that the model learns are not truly general-purpose, though. Human speakers of English who know the meanings of all of the words in a sentence are able to reason about whether that sentence entails any other sentence of English even if they have never seen either sentence before, or have never used the particular pattern of inference before. This broad ability to generalize does not appear to have been learned in these experiments. Pessimistically, the model might be primarily memorizing the reasoning patterns themselves, with any generalization beyond the \textsc{set-out} setting being erratic and case-specific. Optimistically, though, the model might be learning to handle logic in a way more like MacCartney's formal approach, in which the model must know dozens of different possible inference steps involving specific quantifiers and entailment patterns to succeed: if the model hasn't seen the substitution of \ii{most} with \ii{no}, it doesn't have the tools to reason about that substitution. 
%An incomplete ability to generalize is not necessarily detrimental to a model's ability to support reasoning in practice, especially in this later interpretation, since the body of knowledge used to do inference in MacCartney's approach is small enough that it could realistically be learned.

%It remains an open research question whether generalization of the type seen in the \textsc{pair-out} case is possible for any model. A model that could successfully generalize in this way could be potentially valuable both as a more versatile reasoning tool, and as a source of information about the ways in which functional types like quantifiers can be represented in vector space models.
%There are two ways that this problem can be addressed, one practical and one formal.

% An obvious direction for future work is to test for this behavior in more powerful models, or to see whether a dataset containing a more diverse array of reasoning patterns (beyond those possible just with quantification and negation) could encourage generalization. Running experiments like the ones presented above on models with matrices or higher order tensors as word representations \cite{coecke2010mathematical, grefenstette2013towards} might be a promising start.
%It is also not impossible that even an RNTN like the one studied could generalize perfectly: the dramatic instability of the model over the course of training suggests that better approaches to learning could improve performance.

>>>>

Even to the extent that RNTN models can handle functional meanings in the form of the operators of propositional logic, this is not a guarantee that these models can handle functional meanings of the forms seen in natural language. As a first step towards investigating the latter, we attempt to directly measure the degree to which RNNs are able to develop representations for natural language quantifiers like \ii{some} and \ii{all} that are adequate for inference. Quantification is far from the only place in natural language where complex functional meanings are found, but it is a natural starting point, since it can be tested in setences whose structures are otherwise quite simple, and since it has formed a standard case study in prior formal work on natural language inference.

% \subsection{Data}

This experiment replicates similar work described in \citet{bowman2013can}, but replaces the partially manually annotated data in that paper with data that is generated directly using the logical system that we hope to model, yielding results that we believe to be substantially more straightforward to interpret.

@@

% \subsection{Experiments and results}

We evaluate the model using two experimental settings. In the simpler setting we randomly sample 85\% of the data and evaluate on the remaining 15\%. In this setting, the model is being asked to learn a complete reasoning system for the limited language and logic presented in the training data, but it is not being asked to generalize to test examples that are substantially different from those it was trained on. Crucially though, to succeed on this task, the model must be able to recognize all of the lexical relations between the nouns, verbs, and quantifiers and how they interact.

While our primary interest is in discovering the extent to which the model can learn to encode the logic given an arbitrary amount of data, we are also interested in the degree to which the model can infer a correct representation from the logic from more constrained training data. In particular, we propose to segment the sentence pairs according to which quantifiers appear in each pair, and then hold out one such pair for testing. We hypothesize that a model that can efficiently learn to represent a logic should be able to construct an accurate representation of each held out quantifier from the way that it interacts with the other nine quantifiers which are not held out. Since running this experiment requires choosing a pair of quantifiers to hold out before training, the resource demans of training prevent us from testing each of the 55 possible possible pairs of quantifiers, and we choose only four pairs to test on. We attempt choose three pairs of differing quantifiers (\ii{two}/\ii{less-than-two}, \ii{not-all}/\ii{not-most}, and \ii{all}/\ii{some}) to represent six different quantifiers in the test data, choosing pairs that maximize the diversity of relation labels appearing in the test data, and additionally choose one pair in which a quantifier is paired with itself (\ii{no}/\ii{no}).

\begin{table}\small\centering
\begin{tabular}{|l|l|}\hline
\textbf{Training configuration} & \textbf{Test accuracy}\\\hline
\textsc{all split}	&	TODO\% \\\hline
\textsc{pair two/less-than-two}	&	TODO\% \\
\textsc{pair not-all/not-most}	&	TODO\% \\
\textsc{pair all/some}	&	TODO\% \\
\textsc{pair no/no}	&	TODO\% \\
\hline
\end{tabular}

@@

% \subsection{Discussion}

The RNN model was approximately optimal with N dimensional word representations and an M dimensional comparison layer. The RNTN was approximately optimal with N and M dimensions, respectively.
% TODO: Update dimensionality

To be filled in based on Monday's results.\\...\\...\\...\\...\\...\\...\\...\\...\\...\\...

@@

### 2014-06-02 writing/S'14 paper/recursion.tex
Incremental paper update.

Recursive structure is a prominent feature of natural language. Consider, for example, \ii{Alice said hello}, \ii{Bob said that Alice said hello}, and \ii{Carl thinks that Bob said that Alice said hello}. Overt recursion of this kind is easy to find, and theoretical accounts of natural language syntax and semantics rely heavily on recursive structures.
In order for a model to be able to accurately learn natural language meanings, then, we expect that it would need to be able to learn to represent the meanings of function words in a such a way that they are able to behave correctly when taking their own outputs as isput.

We again test this phenomenon within the framework of MacCartney and Manning-style entailment reasoning, but we replace the unanalyzed symbols from the previous experiment with expressions that involve recursive structure. To define these expressions, we turn to propositional logic, a relatively simple logic in which each variable represents either \ii{true} or \ii{false}. We generate data of the form seen in Figure \ref{prop-figure}: strings of arbitrary length consisting of six elementary proposition symbols and the operators \ii{and}, \ii{or}, and \ii{not}, arranged in pairs with the logical relations between them specified. It should be noted that we preserve the same  The data are generated randomly and deduplicated, and consist of about 248k training examples and 44k test examples. Each of the six symbols and each of the three operators is treated as a word for the purposes of our model, and is represented by a randomly initialized vector representation.

% TODO: Worth explicitly calling this project theorem proving?
Socher et al. \cite{socher2012semantic} have previously demonstrated the learning of a logic in a matrix-vector RNN model somewhat similar to our own, but the logic discussed here is substantially stronger, and a much better approximation of the kind of structure that is needed for natural language. The logic learned in that experiment is boolean, wherein the atomic symbols are simply the values 0 and 1, rather than variables over those values. While learning the operators of that logic is not trivial, the ouptuts of each operator can be represented accurately by a single bit. The statements of propositional logic learned here describe conditions on the truth values of propositions where those truth values are not known. As opposed to the two-way contrasts seen in \cite{socher2012semantic}, this logic distinguishes between 64 (2^6) possible assignments of truth values, and expressions of this logic define arbitrary conditions on these possible assignments, for a total of 2^{64} ($\approx 10^{20}$) possible statements that the recursive model needs to be able to distinguish.

@@

>>>>

Recursive structure is a prominent  feature of natural language. Consider, for example, \ii{Alice said hello}, \ii{Bob said that Alice said hello}, and \ii{Carl thinks that Bob said that Alice said hello}. Overt recursion of this kind is easy to find, and theoretical accounts of natural language syntax and semantics rely heavily on recursive structures.
In order for a model to be able to accurately learn natural language meanings, then, we expect that it would need to be able to learn to represent the meanings of function words in a such a way that they are able to behave correctly when taking their own outputs as input.

% Goal: Deep recursion!

We again test this phenomenon within the framework of MacCartney and Manning-style entailment reasoning, but we replace the unanalyzed symbols from the previous experiment with expressions that involve recursive structure. To define these expressions, we turn to propositional logic, a relatively simple logic in which each variable represents either \ii{true} or \ii{false}. We generate data of the form seen in Figure \ref{prop-figure}: strings of arbitrary length consisting of six elementary proposition variables and the operators \ii{and}, \ii{or}, and \ii{not}, arranged in pairs with the logical relations between them specified.

% TODO: Worth explicitly calling this project theorem proving? Yes! There was some confusion at CSLI.

Socher et al. \cite{socher2012semantic} have previously demonstrated the learning of a logic in a matrix-vector RNN model somewhat similar to our own, but the logic discussed here is substantially stronger, and a much better approximation of the kind of structure that is needed for natural language. The logic learned in that experiment is boolean, wherein the atomic symbols are simply the values 0 and 1, rather than variables over those values. While learning the operators of that logic is not trivial, the ouptuts of each operator can be represented accurately by a single bit. The statements of propositional logic learned here describe conditions on the truth values of propositions where those truth values are not known. As opposed to the two-way contrasts seen in \cite{socher2012semantic}, this logic distinguishes between 64 (2^6) possible assignments of truth values, and expressions of this logic define arbitrary conditions on these possible assignments, for a total of 2^{64} ($\approx 10^{20}$) possible statements that the recursive model needs to be able to distinguish. To frame this distinction in another way, the relational statements in our data our theorems about the relations between statements in the logic tested in \cite{socher2012semantic}.

We randomly generate pairs of parentheses-bracketed statements of the logic and then randomly divide the results into training and test data sets. To compute the relation between each pair of statements, we exhaustively enumerate the sets of assignments of truth values to proposition variables that would satisfy each of the statements and then convert the set-theoretic relation between those assignments into one of the seven relations. 
If we do not implement any constraint that the two statements being compared are similar in any way, the generated data consists in large part of statements in which the two refer to largely separate subsets of the six variables, and to which we will nearly always assign the \# relation. In an effort to balance the distribution of relation labels without departing from the basic task of modeling propositional logic, we disallow individual pairs of statements from referring to more than four of the six proposition variables.

We deduplicate the data and discard pairs in which either statement is a tautology or contradiction (a statement that is true of either all or no possible assignments), for which none of the seven relation labels can accurately apply. We then divide the generated pairs into size bins based on the number of logical operators (\ii{and}, \ii{or}, or \ii{not}) in the larger of the two pairs being compared, and discard examples of size greater than twelve by this measure. Finally, we randomly sample 15\% of each bin for a held out test set.
...
65k train

 compute the relaitons between them, discard  , and consist of about 248k training examples and 44k test examples.

 Each of the six symbols and each of the three operators is treated as a word for the purposes of our model, and is represented by a randomly initialized vector representation.

@@

\begin{figure}[t]
\begin{center}
...\\
...\\
...\\
...\\
...\\
...\\
...\\
...\\
...\\
\end{center}

\caption{Model performance by expression size. The vertical dashed line indicates the size of the largest expressions included in the training data.  \label{prop-figure}} 
\end{figure}

The RNN model was approximately optimal with N dimensional word representations and an M dimensional comparison layer. The RNTN was approximately optimal with N and M dimensions, respectively.

% TODO: Include dimensionality
